On the reality of the quantum state once again: A no-go theorem for $\psi$-ontic models

### 2023-02-11 Articles/2021-HarriganSpekkens/HarriganSpekkensTakeTwo.tex
HarriganSpekkens: reviewed with Christine

\begin{abstract}
In 2010 Harrigan and Spekkens (HS) proposed a rigorous classification in order to categorize the nature of the quantum state, i.e.\ to establish whether in a certain model $\psi$ corresponds to a real property of a quantum object, in which case the model is called $\psi$-ontic, or to some observer information, making it $\psi$-epistemic. Building on the HS framework Pusey, Barrett and Rudolph (PBR) proved a theorem claiming that $\psi$-epistemic models contradict quantum mechanics, and thus quantum states cannot be taken to represent mere information about quantum systems. In this essay we propose a simple no-go theorem that explains why also $\psi$-ontic models cannot be expected to reproduce quantum mechanics. We will see that, using information theoretic considerations, the lack of overlap of epistemic states requires all states to be orthogonal, which openly contradicts quantum theory. Combining our result with the PBR argument, our conclusion is that both $\psi$-epistemic and $\psi$-ontic models contradict quantum mechanics.
\end{abstract}

@@

While the discussion has been focused mainly on whether $\psi$-epistemic models are problematic or whether the PBR argument itself is problematic, we ask a different question: what if the HS classification itself is problematic? In particular, the framework assumes that once a full description of the transition probability is given for pure states, mixed states are constructed simply as a classical statistical mixture. But classical mixtures and quantum mixtures have different properties. For example, classical mixtures have a single decomposition in terms of pure states while quantum mixtures do not. Studying the behavior of states under mixing, then, opens a new venue of possible violations that are not as easily seen by looking at transition probability.

The core insight is that the degree of overlap between two probability distribution affects how the entropy of their mixture relates to the entropy of the components. Consequently, we find that the lack of overlap required by $\psi$-ontic models forces all pure states to be orthogonal. Since this openly contradicts quantum theory, we conclude that $\psi$-ontic models are not viable.

Combining our result with the PBR argument, we conclude that the HS classification itself is fundamentally flawed: both $\psi$-epistemic and $\psi$-ontic models do contradict quantum mechanics.\footnote{In principle, PBR leaves open the possibility of $\psi$-epistemic models that violate statistical independence. Given that statistical independence is linked to entropy additivity, that loophole can be closed as well.} Consequently, HS framework should be employed neither to classify interpretations of the quantum formalism, nor to draw conclusion about the nature of $\psi$.

@@

The exposition of the HS classification makes it seem one has complete freedom in choosing the overlap between the overlap between epistemic states. In this section we will show that this is not the case: this is constrained by entropy of mixtures. These constraints cannot be replicated by a $\psi$-ontic model with standard information theoretic techniques. Since quantum statistical mechanics and quantum information theory depend on the correct value of entropy on mixtures, $\psi$-ontic models will violate quantum theory.

To show the violation, we will first study how the entropy behaves when mixing two states, both for the Shannon/Gibbs entropy and the von Neumann entropy.\footnote{Full calculations with all steps are included in the appendix.} The first will be used to calculate entropy of statistical mixtures of epistemic state, since equation \ref{epistemic_mixing} tells us the mixture is done according to a standard (classical) measure of probability. The second will be used to calculate the entropy on the quantum case.

\textbf{Information entropy of mixed non-overlapping probability distributions}. Let $\rho_1$ and $\rho_2$ be two probability density function over a space $\Lambda$ with measure $\lambda$. That is, they are the Radon-Nikodym derivative of the respective probability measures and the measure $\lambda$. The entropy for each distribution is given by the usual formula\footnote{The logarithm is assumed to be in base 2.}, for example
\begin{equation}\label{shannon_entropy}
	H(\rho_1) = - \int_\Lambda \rho_1(\lambda) \log \rho_1(\lambda) d\lambda.
\end{equation}
Suppose the two distributions are disjoint, and let $\rho = \frac{1}{2} \rho_1 + \frac{1}{2} \rho_2$ be a uniform mixture of the two distributions. More precisely, $\rho$ is the Radon-Nikodym derivative of the average of the two probability measures with respect to $\lambda$. The entropy of $\rho$ is given by
\begin{equation}\label{entropy_nonoverlap}
	H(\rho) = 1 + \frac{1}{2} H(\rho_1) + \frac{1}{2} H(\rho_2).
\end{equation}
Note how the non-overlapping assumption fixes the entropy of the mixed state.

\textbf{Quantum information entropy of quantum mixed states}. Now suppose $\psi$ and $\phi$ are two pure quantum states and let $p = | \langle \psi | \phi \rangle |^2$ be the probability of transition from one to the other. Consider the mixed state $\rho = \frac{1}{2} | \psi \rangle \langle \psi | + \frac{1}{2} | \phi \rangle \langle \phi |$. Its entropy is given by
\begin{equation}\label{entropy_mixed}
	H(\rho) = H\left(\frac{1+\sqrt{p}}{2}, \frac{1-\sqrt{p}}{2}\right).
\end{equation}
where the right side of the equation is the Shannon entropy calculated on the given values. Therefore the entropy of an equal mixture of two pure states depends only on the probability of measuring one having prepared the other.

@@

\emph{Proof}: Suppose we have a $\psi$-ontic model defined according to \cite{Harrigan:2010}. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping probability distributions over a space $\Lambda$. That is, they are the two probability density functions defined to be the Radon-Nikodym derivatives between the respective probability measures over $\Lambda$ and the measure $\lambda$. Equation \ref{epistemic_mixing} tells us that the probability measure combine in the usual way, and therefore an equal mixture of the two states must obey \ref{entropy_nonoverlap}. Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states in quantum mechanics is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
		H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
		H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
		&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq.~\ref{entropy_mixed}, it follows that $p$ must be zero. In fact, the only case the Shannon entropy for a two element distribution is equal to one is for the uniform distribution $H(1/2, 1/2) = 1$.  That is, we must have that
\begin{equation}\label{orthogonal}
	\langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

@@

If we combine this result with PBR, the only models that are not ruled our are $\psi$-epistemic models where the objects are not independently prepared. That is, where the joint distribution over ontic states is not the product of the marginals. However, this loophole can also been closed with information theoretic considerations. It is a well-known result in both classical and quantum information theory that the entropy of a joint distribution is the sum of the marginal if and only if the subsystems are independent (\cite{Ash:2010}, \cite{Nielsen:2010}). Therefore an epistemic state that represents independent quantum states must also be the joint distribution of statistically independent epistemic states of the individual systems. This tells us that $\psi$-epistemic models must represent composite states of independently prepared systems with independently distributed epistemic states, and therefore PBR applies. This closes the last loophole: no model in the HS categorization can reproduce quantum theory in its entirety, which includes quantum information and quantum statistical mechanics.

Thus, one concludes the following:
\begin{quote}
	\textbf{no ontological model can reproduce quantum statistical mechanics and quantum information with standard information theoretic tools}.
\end{quote}
That is, even if one had an ontological model that mapped the result of single outcome correctly, it would still give the wrong answer when trying to calculate ensembles and their entropy.

@@

First of all, information entropy is strictly concave, meaning that $H(p \rho_1 + (1-p) \rho_2) \geq p H(\rho_1) + (1-p) H(\rho_2)$ with the equality valid only if $\rho_1$ and $\rho_2$ are the same exact state. That is, we can check whether two states $\rho_1$ and $\rho_2$ are the same state simply by creating a mixture $\rho = \frac{1}{2} \rho_1 + \frac{1}{2} \rho_2$ and see whether we end up with the same entropy or not. In other words: \emph{entropy fixes equality}. %spiega meglio
The issue here is that the mapping between epistemic states and density matrices is not injective. Quantum mixtures, in fact, do not have a single decomposition in pure states. For example, the maximally mixed state of a qubit can be achieved with either an equal mixture of spin up and down or an equal mixture of spin left and right. Therefore the same quantum mixture must correspond to an equivalence class of epistemic states, which means whatever definition we have on entropy for epistemic states must not satisfy strict concavity to be able to agree with the quantum case.\footnote{The degree of degeneracy one has to require is also extreme. For a qubit, in the simplest case, we would have a $\psi$-ontic state for each direction of spin, meaning the space of ontic states is the sphere $S^2$. The space of all possible epistemic states is the set of all possible distribution integrable to one. This is an infinite dimensional function space. However, the space of mixed state for a quibit is the interior of a sphere, which is a bounded three-dimensional manifold. The equivalence classes of epistic states, then, must be infinite dimensional.} That is, one can make mixtures of different ensemble without increasing entropy, which does not make physical sense.

Secondly, as equation \ref{entropy_mixed} shows, computing the entropy requires the inner product. In fact, if we had the entropy for all even mixtures of all pairs of pure states, we could reconstruct the inner product. Therefore, whatever other structure one puts on top of epistemic states, essentially redefines the inner product. Also, it has to do it in a way compatible with the other definitions of the model, therefore this structure will likely be ad-hoc for each model. In other words, the idea that the inner product just represents transition probabilities, which, in our view, is the key idea the ontological model uses to explain ``what really happens'' does not work.

@@

The key problem of the HS categorization is that it is ``not quantum enough''. The framework tries to hide the non-classicality at the level of each ontic state, which, by construction, cannot be decomposed further using a standard probability measure. However, it posits that statistical mixtures are simply classical mixture in the sense that they are captured by a standard probability measure. But quantum mixtures are non-classical in the same way that the quantum transition probabilities defined by the inner product are non-classical. Therefore the HS categorization prevents one to describe quantum statistical mixing.

Our conclusion is that the PBR theorem finds that $\psi$-epistemic models are not sustainable not because there is a problem with $\psi$-epistemic models in particular, but rather there is an underlying problem in the HS definitions. If understood in this manner, the PBR result is the first part of a more general argument showing the inadequacy of the HS framework to categorize quantum theories. Argument that we have completed with the proposed no-go theorem against $\psi$-ontic models.

>>>>

\begin{abstract}
	In this paper we show that $\psi$-ontic models, as defined by Harrigan and Spekkens (HS), cannot reproduce quantum theory. Instead of focusing on probability, we use information theoretic considerations to show that all pure states of $\psi$-ontic models must be orthogonal to each other, in clear violation of quantum mechanics. Given that  (1) Pusey, Barrett and Rudolph (PBR) previously showed that $\psi$-epistemic models, as defined by HS, also contradict quantum mechanics, and (2) the HS categorization is exhausted by these two types of models, we conclude that the HS categorization itself is problematic as it leaves no space for models that can reproduce quantum theory.
\end{abstract}

@@

While the discussion has been focused mainly on whether $\psi$-epistemic models are problematic or whether the PBR argument itself is problematic, we ask a different question: what if the underlying HS classification itself is problematic? In particular, the framework assumes that once pure states are fully characterized, mixed states are constructed simply as a standard (i.e. classical) statistical mixture. But classical mixtures and quantum mixtures have different properties. For example, classical mixtures have a single decomposition in terms of pure states while quantum mixtures do not. Studying the behavior of states under mixing, then, provides a new way to explore possible violations.

Using this new approach, in this paper we are going to provide a no-go theorem for $\psi$-ontic models. The core insight is that the degree of overlap between two probability distributions affects how the entropy of their mixture relates to the entropy of the components. Consequently, we find that the lack of overlap required by $\psi$-ontic models forces all pure states to be orthogonal. Since this directly contradicts quantum theory, we conclude that $\psi$-ontic models are not viable.

Combining our result with the PBR argument, we conclude that the HS classification itself is fundamentally flawed: both $\psi$-epistemic and $\psi$-ontic models contradict quantum mechanics.\footnote{In principle, PBR leaves open the possibility of $\psi$-epistemic models that violate statistical independence. Given that statistical independence is linked to entropy additivity, that loophole can be closed as well.} Consequently, the HS framework should be employed neither to classify interpretations of the quantum formalism, nor to draw conclusions about the nature of $\psi$.

@@

The exposition of the HS classification makes it seem one has complete freedom in choosing the overlap between epistemic states. In this section we will show that this is not the case: this is constrained by entropy of mixtures. These constraints cannot be replicated by a $\psi$-ontic model with standard information theoretic techniques. Since quantum statistical mechanics and quantum information theory depend on the correct value of entropy on mixtures, $\psi$-ontic models will violate quantum theory.

To show the violation, we will first study how the entropy behaves when mixing two states, both for the Shannon/Gibbs entropy and the von Neumann entropy.\footnote{Full calculations with all steps are included in the appendix.} The first will be used to calculate entropy of statistical mixtures of epistemic states, since equation \ref{epistemic_mixing} tells us the mixture is done according to a standard (classical) measure of probability. The second will be used to calculate the entropy on the quantum case.

\textbf{Information entropy of mixed non-overlapping probability distributions}. Let $\rho_1$ and $\rho_2$ be two probability density functions over a space $\Lambda$ with measure $\lambda$.\footnote{The notation of HS uses $\lambda$ as both the value and measure for integration, as often done in single variable integrals.} That is, they are the Radon-Nikodym derivatives of the respective probability measures with respect to the measure $\lambda$. The entropy for each distribution is given by the usual formula\footnote{The logarithm is assumed to be in base 2.}, for example
\begin{equation}\label{shannon_entropy}
	H(\rho_1) = - \int_\Lambda \rho_1(\lambda) \log \rho_1(\lambda) d\lambda.
\end{equation}
Suppose the two distributions are disjoint, and let $\rho = \frac{1}{2} \rho_1 + \frac{1}{2} \rho_2$ be a uniform mixture of the two distributions. More precisely, $\rho$ is the Radon-Nikodym derivative of the average of the two probability measures with respect to $\lambda$. The entropy of $\rho$ is given by
\begin{equation}\label{entropy_nonoverlap}
	H(\rho) = 1 + \frac{1}{2} H(\rho_1) + \frac{1}{2} H(\rho_2).
\end{equation}
Note how the non-overlapping assumption fixes the entropy of the mixed state.

\textbf{Quantum information entropy of quantum mixed states}. Now suppose $\psi$ and $\phi$ are two pure quantum states and let $p = | \langle \psi | \phi \rangle |^2$ be the probability of transition from one to the other. Consider the mixed state $\rho = \frac{1}{2} | \psi \rangle \langle \psi | + \frac{1}{2} | \phi \rangle \langle \phi |$. Its entropy is given by
\begin{equation}\label{entropy_mixed}
	\begin{aligned}
	H(\rho) &= H\left(\frac{1+\sqrt{p}}{2}, \frac{1-\sqrt{p}}{2}\right) \\
	&= - \frac{1+\sqrt{p}}{2} \log \frac{1+\sqrt{p}}{2} - \frac{1-\sqrt{p}}{2} \log \frac{1-\sqrt{p}}{2} ,
	\end{aligned}
\end{equation}
where the right side of the equation is the Shannon entropy calculated on the given values. Therefore the entropy of an equal mixture of two pure states depends only on the probability of measuring one having prepared the other.

@@

\emph{Proof}: Suppose we have a $\psi$-ontic model defined according to \cite{Harrigan:2010}. The epistemic states $p(\lambda|P_\psi)$ and $p(\lambda|P_\phi)$ consist of non-overlapping probability distributions over a space $\Lambda$. That is, they are two non-overlapping probability density functions defined to be the Radon-Nikodym derivatives between the respective probability measures over $\Lambda$ and the measure $\lambda$. Equation \ref{epistemic_mixing} tells us that the probability measures combine in the usual way, and therefore an equal mixture of the two states must obey equation \ref{entropy_nonoverlap}. Given that $\psi$ and $\phi$ are pure states, and the entropy for pure states in quantum mechanics is zero, we must have
\begin{equation}\label{entropy_pure}
	H(p(\lambda|P_\psi)) = H(p(\lambda|P_\phi)) = 0,
\end{equation}
and therefore
\begin{equation}\label{required_entropy}
	\begin{aligned}
		H\left(p(\lambda|\frac{1}{2}P_\psi + \frac{1}{2}P_\phi)\right) &= \\
		H\left(\frac{1}{2}p(\lambda|P_\psi) + \frac{1}{2}p(\lambda|P_\phi)\right) 
		&= 1.
	\end{aligned}
\end{equation}
If we compare the above with eq.~\ref{entropy_mixed}, it follows that $p$ must be zero. In fact, the only case in which the Shannon entropy for a two-element distribution is equal to one is for the uniform distribution $H(1/2, 1/2) = 1$.  That is, we must have that
\begin{equation}\label{orthogonal}
	\langle \psi | \phi \rangle = 0
\end{equation}
no matter what $\psi$ and $\phi$ are.

@@

If we combine this result with PBR, the only models that are not ruled out are $\psi$-epistemic models where the objects are not independently prepared. That is, where the joint distribution over ontic states is not the product of the marginals. However, this loophole can also be closed with information theoretic considerations. It is a well-known result in both classical and quantum information theory that the entropy of a joint distribution is the sum of the marginal if and only if the subsystems are independent (\cite{Ash:2010}, \cite{Nielsen:2010}). Therefore an epistemic state that represents independent quantum states must also be the joint distribution of statistically independent epistemic states of the individual systems. This tells us that $\psi$-epistemic models must represent composite states of independently prepared systems with independently distributed epistemic states, and therefore PBR applies. This closes the last loophole: no model in the HS categorization can reproduce quantum theory in its entirety, which includes quantum information and quantum statistical mechanics.

Thus, one concludes the following:
\begin{quote}
	\textbf{No ontological model can reproduce quantum theory}.
\end{quote}
Note that HS do not claim that all interpretations must adhere to their framework, stating (\cite{Harrigan:2010}, p. 134), ``Therefore, to categorize any given interpretation of the quantum formalism as $\psi$-ontic or $\psi$-epistemic, it is first necessary to cast it into the mold of an ontological model. If an interpretation resists being so cast, then it cannot be fit into our categorization.'' The result is that a successful interpretation of quantum mechanics must resists being so cast.

@@

First of all, information entropy is strictly concave, meaning that $H(p \rho_1 + (1-p) \rho_2) \geq p H(\rho_1) + (1-p) H(\rho_2)$ with the equality valid only if $\rho_1$ and $\rho_2$ are the same exact state. That is, we can check whether two states $\rho_1$ and $\rho_2$ are the same state simply by creating a mixture $\rho = \frac{1}{2} \rho_1 + \frac{1}{2} \rho_2$ and see whether we end up with the same entropy or not. In other words: \emph{entropy fixes equality}. %spiega meglio
The issue here is that the mapping between epistemic states and density matrices is not injective. Quantum mixtures, in fact, do not have a single decomposition in pure states. For example, the maximally mixed state of a qubit for a spin-$1/2$ system can be achieved with either an equal mixture of spin up and down or an equal mixture of spin left and right. Therefore epistemic states cannot be mapped one-to-one to quantum density operators, but rather each density operator corresponds to an equivalence class of epistemic states.\footnote{The degree of degeneracy one has to require is also extreme. For a qubit, in the simplest case, we would have a $\psi$-ontic state for each direction on the Bloch sphere, meaning the space of ontic states is the sphere $S^2$. The space of all possible epistemic states is the set of all possible distributions integrable to one. This is an infinite dimensional function space. However, the space of mixed states for a quibit is the interior of the Bloch sphere, which is a bounded three-dimensional manifold. The equivalence classes of epistemic states, then, must be infinite dimensional.} This degeneracy means that the notion of equality among epistemic states is different than the notion of equality among density operators. In turn, this means that whatever definition we have for entropy on epistemic states must not satisfy strict concavity, if we want the value of entropy to agree between epistemic states and density operators. \emph{Entropy cannot fix equality for epistemic states}. In other words, one can make mixtures of different epistemic states without increasing entropy, which is questionable on physics grounds.

Secondly, as equation \ref{entropy_mixed} shows, computing the entropy requires the inner product. In fact, if we had the entropy for all uniform mixtures of all pairs of pure states, we could reconstruct the inner product. Therefore, whatever other structure one puts on top of epistemic states essentially redefines the inner product. Also, it has to do it in a way compatible with the other definitions of the model, therefore this structure will likely be ad-hoc for each model. In other words, the idea that the inner product just represents transition probabilities, which, in our view, is the key idea the ontological model uses to explain ``what really happens'' does not work.

@@

The key problem of the HS categorization is that it is ``not quantum enough.'' The framework tries to hide the non-classicality at the level of each ontic state, which, by construction, cannot be decomposed further using a standard probability measure. However, it posits that statistical mixtures are simply classical mixtures in the sense that they are captured by a standard probability measure. But quantum mixtures are non-classical in the same way that the quantum transition probabilities defined by the inner product are non-classical. Therefore the HS categorization prevents one from describing quantum statistical mixing.

Our conclusion is that the PBR theorem finds that $\psi$-epistemic models are not sustainable not because there is a problem with $\psi$-epistemic models in particular, but rather there is an underlying problem in the HS definitions. If understood in this manner, the PBR result is the first part of a more general argument showing the inadequacy of the HS framework to categorize quantum theories, an argument that we have completed with the proposed no-go theorem against $\psi$-ontic models.
